---
title: "Data Product- America’s unique gun violence problem"
author: "Preethi S Ranganathan"
date: "06/02/2018"
output: github_document
---

**Data Product:** America’s unique gun violence problem
https://www.vox.com/policy-and-politics/2017/10/2/16399418/us-gun-violence-statistics-maps-charts

The data product, is about the Gun Violence in the United States. The maps and charts in the data product show what gun violence looks like compared with the rest of the world and also how prevalent it is in the US, why it happens, and why it’s such a tough problem to fix.

There are two objectives with this project:

1. Redesign product - Replication and Redesign of the original data product. 
2. Deceptive product - Development of deceptive data product, which explains intentional and unintentional distortions to the data.

There are three claims in each product, which are discussed in detail below:

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(gdata)
library(gridExtra)
library(stringr)
library(varhandle)
library(ggpubr)
library(data.table)
library(png)
```

## Redesign product:

### Claim 1: United States has the highest firearm homicide and suicide rates among the top most developed countries.

#### Chart from Vox:
```{r Original Claim 1}
img <- readPNG( '/Users/preethiranganathan/Documents/Santa Clara/data viz/indiv proj/Image4.png',native=TRUE)
plot.new() 
rasterImage(img,0,0,1,1)
```

<sup>Source: Vox https://www.vox.com/policy-and-politics/2017/10/2/16399418/us-gun-violence-statistics-maps-charts </sup>

This chart tells us about the mass shootings, which make up a tiny portion of America’s firearm deaths, since 2012.

#### First Redesign version:

The first redesign version is a replication of the original chart in the data product with the only change being the number of casualties represented by both color and size of circles. This change was made to show the severity of shootings more clearly.

```{r Redesign First Claim 1}
img <- readPNG( '/Users/preethiranganathan/Documents/Santa Clara/data viz/indiv proj/Image3.png',native=TRUE)
plot.new() 
rasterImage(img,0,0,1,1)
```

<sup>Source: Gun Violence Archive: http://www.gunviolencearchive.org </sup>

#### Final Redesign version:

The main intent of this graph is to show the severity and importance of gun violence in the US. The first claim 'There has been a lot of mass shootings since 2012' do not reflect that point, precisely. Hence, in the revised version, I altered the claim to indicate the importance of gun violence.

The altered claim is 'United States has the highest firearm homicide and suicide rates among the top most developed countries'. This claim would convey the message of why firearm violence needs to be addressed immediately.

The other point is that the original chart is a geographical map. But it is not a good idea to have this kind of map as a warrant for expressing number of casualties. It shows that gun casualties occur at places with high population.

For coming up with this chart, I made a comparison of homicide and suicides rates of US with other developed countries to show that United States has the highest firearm homicide and suicide rates among the top most developed countries.

The developed countries list is from https://www.topteny.com/top-10-highly-developed-countries-in-the-world/

```{r Redesign Final Claim 1}
ghom= read.csv('/Users/preethiranganathan/Documents/Santa Clara/data viz/indiv proj/homicide_rate_global.csv',header=TRUE)

ghom1=ghom[ghom$Location %in% c('Canada','Germany' ,'United States','Netherlands','Australia','United Kingdom','Norway','New Zealand', 'Switzerland'),] 

p1<-ggplot(ghom1, aes(reorder(ghom1$Location, ghom1$Value),Value)) +geom_col(position ='dodge') + labs(y = "Firearm Homicides per 100,000 people",x='')+ coord_flip()+theme_minimal()+ggtitle('US has the highest firearm homicide and suicide rates among the top most developed countries')+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),axis.line = element_line(colour = "black"), plot.title = element_text( size=9),axis.title= element_text( size=8))

gsuicide= read.csv('/Users/preethiranganathan/Documents/Santa Clara/data viz/indiv proj/suicide_rate_global.csv',header=TRUE)
gsuicide1=gsuicide[gsuicide$Location %in% c('Canada','Germany' ,'United States', 'Netherlands','Australia','United Kingdom','Norway','New Zealand','Switzerland'),] 

p2<-ggplot(gsuicide1, aes(reorder(gsuicide1$Location, gsuicide1$Value),Value)) +geom_col(position ='dodge') + labs(y = "Firearm Suicides per 100,000 people",x='')+ coord_flip()+theme_minimal()+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),axis.line = element_line(colour = "black"), plot.title = element_text( size=11),axis.title= element_text( size=8))

grid.arrange(p1,p2,nrow=2)

```

<sup>Source: Institute for Health Metrics and Evaluation: https://vizhub.healthdata.org/gbd-compare/</sup>

This dataset has firearm-related homicide rates for all countries for the year 2016 and I have picked only few developed countries for comparison as it is only fair to compare death rates of countries with similiar living standards, given by Human Development Index. 

Use of colors is not necessary here because the bars itself gives a good comparison of death rates for different countries.


### Claim 2: Firearm suicide rates has been increasing in the US over years.

#### Chart from Vox:
```{r Original Claim 2}
img <- readPNG( '/Users/preethiranganathan/Documents/Santa Clara/data viz/indiv proj/Image 1.png',native=TRUE)
plot.new() 
rasterImage(img,0,0,1,1)
```

<sup>Source: Vox https://www.vox.com/policy-and-politics/2017/10/2/16399418/us-gun-violence-statistics-maps-charts</sup>

Chart 10 in Vox claims that most firearm deaths are suicides. In this chart, we see a time series of firearm suicides and homicides from 1999 to 2016 and we can see that number of firearm homicides is always less than the number of firearm suicides over years.


#### First Redesign version:

I made two charts for the first version, first one similiar to the vox chart except that the number of deaths is now normalized per 100,000 people. It still shows the same homicide and suicide trend.

The second chart, shows the suicides as a perecentage of firearm deaths. I thought it is a good idea, because the claim talks about suicides as a proportion of firearm deaths, which is what the claim 'Most gun deaths are suicides' tries to convey.

```{r Redesign First Claim 2}
img <- readPNG( '/Users/preethiranganathan/Documents/Santa Clara/data viz/indiv proj/Image5.png',native=TRUE)
plot.new() 
rasterImage(img,0,0,1,1)
```

<sup> Source: Centers for Disease Control and Prevention: https://webappa.cdc.gov/sasweb/ncipc/mortrate10_us.html </sup>

Note that the data pixel ratio is very low in the above chart, as only the top of the bars convey information. This will be improved in the final version.

```{r Redesign First Claim 2 cont..}
img <- readPNG( '/Users/preethiranganathan/Documents/Santa Clara/data viz/indiv proj/Image7.png',native=TRUE)
plot.new() 
rasterImage(img,0,0,1,1)
```

<sup> Source: Centers for Disease Control and Prevention: https://webappa.cdc.gov/sasweb/ncipc/mortrate10_us.html </sup>

#### Final Redesign version:

The point that I wanted to make with the claim is that firearm suicides, being the major proportion of firearm deaths are increasing over the years.  

First, I designed a chart to convey the message of how firearm suicide rates are increasing in the US over years and then the second chart to explain why we have to address suicides immediately.

The dataset used here has raw number of suicides every year from 1999 to 2016. It makes more sense to use normalized suicide rates as metric, because using raw numbers would not exclude the population factor into account, as it keeps increasing over years.

```{r Redesign Final Claim 2}
#Firearm Suicide data files from 1999 to 2015
suicidefiles <- list.files(path='~/Documents/Santa Clara/data viz/indiv proj/firearm_suicide',full.names = TRUE, pattern = ".csv")
d=data.frame()
for(i in 1:length(suicidefiles)){
  d=rbind(read.table(suicidefiles[i],sep=',',header=TRUE),d)
}
suicide=data.frame(d[,6],d[,8],d[,9],d[,10])
suicide=unique(suicide)
colnames(suicide)<- c('Year','Suicide_Method','Deaths','Population')
#Firearm Suicides per 100K
suicide['Num_of_Deaths_in_100K']=(suicide['Deaths']/suicide['Population'])*100000
suicide=suicide[sort(suicide$Year,decreasing=FALSE,index.return=TRUE)$ix,]

ggplot(suicide) +geom_line(aes(Year, Num_of_Deaths_in_100K ))+ggtitle('Firearm suicide rates has been increasing in the US over years')+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),  axis.line = element_line(colour = "black"), plot.title = element_text( size=11), axis.title= element_text( size=9))+ labs(x='Year',y = "Number of suicides per 100,000 people") + scale_x_continuous(breaks=seq(1999,2016,1))
```

<sup> Source: Centers for Disease Control and Prevention (https://webappa.cdc.gov/sasweb/ncipc/mortrate10_us.html) </sup>

The number of firearm suicides was gradually dropping from 2002 to 2006, but it started increasing rapidly since 2007. This could be because of global financial crisis with high numbers of people killing themselves suffering heavy job losses. 
https://journalistsresource.org/studies/society/public-health/suicides-during-great-recession-united-states-canada-europe

Added to this, people pre-dominantly choose firearm for killing themselves compared to any other methods in the US.
```{r Redesign Final Claim 2 cont..}
#Poisoning Suicide data files from 1999 to 2015
psuicidefiles <- list.files(path='~/Documents/Santa Clara/data viz/indiv proj/poisoning_suicide',full.names = TRUE, pattern = ".csv")
d=data.frame()
for(i in 1:length(psuicidefiles)){
  d=rbind(read.table(psuicidefiles[i],sep=',',header=TRUE),d)
}
psuicide=data.frame(d[,6],d[,8],d[,9],d[,10])
psuicide=unique(psuicide)
colnames(psuicide)<- c('Year','Suicide_Method','Deaths','Population')
psuicide['Suicide_Method']='Poisioning'

#Poisoning Suicides per 100K
psuicide['Num_of_Deaths_in_100K']=(psuicide['Deaths']/psuicide['Population'])*100000

#Cut Suicide data files from 1999 to 2015
csuicidefiles <- list.files(path='~/Documents/Santa Clara/data viz/indiv proj/cut_suicide',full.names = TRUE, pattern = ".csv")
d=data.frame()
for(i in 1:length(psuicidefiles)){
  d=rbind(read.table(csuicidefiles[i],sep=',',header=TRUE),d)
}
csuicide=data.frame(d[,6],d[,8],d[,9],d[,10])
csuicide=unique(csuicide)
colnames(csuicide)<- c('Year','Suicide_Method','Deaths','Population')
csuicide['Suicide_Method']='Cut/Pierce'

#Cut Suicides per 100K
csuicide['Num_of_Deaths_in_100K']=(csuicide['Deaths']/csuicide['Population'])*100000

#Suffocation Suicide data files from 1999 to 2015
ssuicidefiles <- list.files(path='~/Documents/Santa Clara/data viz/indiv proj/suffocation_suicide',full.names = TRUE, pattern = ".csv")
d=data.frame()
for(i in 1:length(ssuicidefiles)){
  d=rbind(read.table(ssuicidefiles[i],sep=',',header=TRUE),d)
}
ssuicide=data.frame(d[,6],d[,8],d[,9],d[,10])
ssuicide=unique(ssuicide)
colnames(ssuicide)<- c('Year','Suicide_Method','Deaths','Population')
#Suffocation Suicides per 100K
ssuicide['Num_of_Deaths_in_100K']=(ssuicide['Deaths']/ssuicide['Population'])*100000
ssuicide['Suicide_Method']='Suffocation'
suicide['Suicide_Method']='Firearm'

#Combining rows of homicide and suicide data frame.
totalsuicides=rbind(suicide,csuicide,psuicide,ssuicide)
compressed <- totalsuicides %>% group_by(Suicide_Method) %>% summarize(total=sum(Deaths,na.rm=TRUE))

slices= compressed$total
lbls =compressed$Suicide_Method
pie(slices, labels = lbls, main="Firearm suicides form the major proportion of suicides in the US",col=c('#B79F00','#00BFC4','orange','#F8766D'),font.main=1,cex.main =0.9)

```

<sup> Source: Centers for Disease Control and Prevention: https://webappa.cdc.gov/sasweb/ncipc/mortrate10_us.html </sup>

Contrasting colors are used here to just show difference between propotion of suicide methods and it doesn't convey any meaning, in particular. 

Note: Throughout the final version charts, the above colors will be used uniformly (only if colors are needed).

### Claim 3: States with the most guns report the most firearm homicides and suicides.

#### Chart from Vox:
```{r Original Claim 3}
img <- readPNG( '/Users/preethiranganathan/Documents/Santa Clara/data viz/indiv proj/Image 2.png',native=TRUE)
plot.new() 
rasterImage(img,0,0,1,1)
```

<sup>Source: Vox https://www.vox.com/policy-and-politics/2017/10/2/16399418/us-gun-violence-statistics-maps-charts</sup>

The claim in the original chart states that more guns, more sucides. Different colors are used to compare between firearm and non-firearm suicides. The left-most bars indicate states with high gun ownership and the right-most bars indicate states with low gun ownership.

#### First Redesign Version:

While redesigning the chart, I made a change to present two side-by-side charts, as it takes time to understand what highest and lowest rates mean.

Here, we have grouped top 10 states with high gun ownership rate into one group and top 10 states with low gun ownership rate into anothe group. This makes the warrant more clear to support the claim 'more guns more suicides'.

Red color is used here as I connected red to violence. This assumption is far-fetched. Not all audience will make the same connection. Hence, it is better to go with black and white, as colors doesn't convey anything here. This will be improved in the final version chart.

```{r Redesign First Claim 3}
img <- readPNG( '/Users/preethiranganathan/Documents/Santa Clara/data viz/indiv proj/Image 6.png',native=TRUE)
plot.new() 
rasterImage(img,0,0,1,1)
```

<sup>Sources: N.C. State Center for Health Statistics- http://www.washingtonpost.com/wp-srv/health/interactives/guns/ownership.html?noredirect=on,<br /> Gun control laws - https://www.statefirearmlaws.org/table.html</sup>

#### Final Redesign Version:

In the first version, only firearm suicides were compared in states with different levels of gun ownership rates. Hence, including homcides rates for different gun ownership levels in the final version. And, there is no point of showing non-firearm suicide rates, as the intent is to compare firearm rates between states with high and low gun ownership rates. 

```{r Redesigned Final Claim 3}
#gun ownership survey data
df = read.xls ("/Users/preethiranganathan/Documents/Santa Clara/data viz/indiv proj/gun ownership statewise.xlsx", sheet = 1, header = TRUE)
gun_ownership=data.frame(df[,1],df[,2],df[,3],df[,4])
colnames(gun_ownership)<-c('State','Number_surveyed','Num.gunowned','Num.gunowned.percent')

# making dataframe for states with less gun ownership rate
low=data.frame(gun_ownership[order(gun_ownership$Num.gunowned.percent),'State'][1:10])
low$Ownershiprate='States with low gun ownership rate'

# making dataframe for states with high gun ownership rate
high=data.frame(gun_ownership[order(-gun_ownership$Num.gunowned.percent),'State'][1:10])
high$Ownershiprate='States with high gun ownership rate'
colnames(high)[1]<-'State'
colnames(low)[1]<-'State'

# combining rows of states with high and low gun ownership rates into a dataframe
highlow<-rbind(high,low)
highlow$State<-str_replace_all(highlow$State,"[*]","")

# statewise firearm suicides
files <- list.files(path='/Users/preethiranganathan/Documents/Santa Clara/data viz/indiv proj/statewise_gunsuicide',full.names = TRUE, pattern = ".csv")
d=data.frame()
for(i in 1:length(files)){
  d=rbind(read.table(files[i],sep=',',header=TRUE),d)
}
state_suicide=data.frame(d[,3],d[,9],d[,10])
state_suicide=unique(state_suicide)
colnames(state_suicide)=c('State','Deaths','Population')
# statewise firearm suicides per 100K
state_suicide$Deathsper100K=state_suicide$Deaths/state_suicide$Population*100000
state_suicide$Death_type ='Suicides'


# statewise firearm homicides
files <- list.files(path='/Users/preethiranganathan/Documents/Santa Clara/data viz/indiv proj/statewise_gunhomicide',full.names = TRUE, pattern = ".csv")
d=data.frame()
for(i in 1:length(files)){
  d=rbind(read.table(files[i],sep=',',header=TRUE),d)
}
state_homicide=data.frame(d[,3],d[,9],d[,10])
state_homicide=unique(state_homicide)
colnames(state_homicide)=c('State','Deaths','Population')

# statewise non-firearm suicides per 100K
state_homicide$Deathsper100K=state_homicide$Deaths/state_homicide$Population*100000
state_homicide$Death_type ='Homicides'

state_deaths=rbind(state_homicide,state_suicide)

ownership_deaths=inner_join(state_deaths,highlow,by='State')

ggplot(ownership_deaths, aes(Ownershiprate, Deathsper100K, fill=Death_type)) +geom_col(position ='dodge')+ylim(0,14)+ ggtitle('States with the most guns report the most firearm homicides and suicides') + labs(x='',y = "Deaths per 100,000 people")+ theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),  axis.line = element_line(colour = "black"), plot.title = element_text( size=10), axis.title= element_text( size=8), legend.title= element_text(size=8))+ scale_fill_manual(name = "Death Type",values = c("#F8766D", "#00BFC4"))

```

<sup>Sources: N.C. State Center for Health Statistics- http://www.washingtonpost.com/wp-srv/health/interactives/guns/ownership.html?noredirect=on, Gun control laws - https://www.statefirearmlaws.org/table.html</sup>

This chart clearly shows that death rates are higher for states with high gun ownership rate compared to states with low gun ownership rates.

Contrasting colors are used here to just show comparison and it doesn't convey any meaning, in particular.

## Deceptive product:

### Deceptive Claim 1: Homicide rates have declined over years while the firearm homicide rates have not (there is no increasing/decreasing trend over years).

#### Actual Argument:
Firearm and non-firearm homicide rates have declined over the years.(Chart 9) 

#### Actual Plot (1981-2016):

```{r Actual DClaim 1, message=FALSE, warning=FALSE }
#Firearm Homicide data files from 1981 to 2015
firearmhomicidefiles <- list.files(path='~/Documents/Santa Clara/data viz/indiv proj/homicide',full.names = TRUE, pattern = ".csv")
d=data.frame()
for(i in 1:length(firearmhomicidefiles)){
  d=rbind(read.table(firearmhomicidefiles[i],sep=',',header=TRUE),d)
}
firearm_homicide=data.frame(d[,6],d[,9],d[,10])
firearm_homicide=unique(firearm_homicide)
colnames(firearm_homicide)<- c('Year','Deaths','Population')
firearm_homicide$Homicide_type='Firearm'

#Non-Firearm Homicide data files from 1981 to 2015
nonfirearmhomicidefiles <- list.files(path='~/Documents/Santa Clara/data viz/indiv proj/non-firearm-homicides',full.names = TRUE, pattern = ".csv")
d=data.frame()
for(i in 1:length(nonfirearmhomicidefiles)){
  d=rbind(read.table(nonfirearmhomicidefiles[i],sep=',',header=TRUE),d)
}
nonfirearm_homicides=data.frame(d[,7],d[,9],d[,10])
nonfirearm_homicides=unique(nonfirearm_homicides)
colnames(nonfirearm_homicides)<- c('Year','Deaths','Population')
nonfirearm_homicides$Homicide_type='Non-Firearm'

#Merge and normalized
homicides=merge(firearm_homicide, nonfirearm_homicides ,by.x='Year',by.y='Year',all=TRUE)
homicides=data.frame(homicides[,1],homicides[,2],homicides[,3],homicides[,4])
colnames(homicides)<-c('Year','Firearm','Population','Nonfirearm')
homicides$Firearm_norm<-(homicides$Firearm/homicides$Population)*100000
homicides$NonFirearm_norm<-(homicides$Nonfirearm/homicides$Population)*100000

#Combining firearm and non-firearm homicides
homicides1=rbind(firearm_homicide,nonfirearm_homicides)
homicides1$norm<-(homicides1$Deaths/homicides1$Population)*100000

ggplot(homicides1) +geom_line(aes(Year, norm,color=Homicide_type ))+ggtitle('Firearm and Non-Firearm homicide rates have declined over the years')+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black"), plot.title = element_text( size=10),axis.title= element_text( size=9),legend.title= element_text( size=9))+ labs(x='Year',y = "Number of homicides per 100,000 people")+ scale_x_continuous(breaks=seq(1981,2017,3))+ coord_fixed(ratio = 4)+ guides(col=guide_legend(title="Homicide type"))
```

<sup>Source: Centers for Disease Control and Prevention -https://webappa.cdc.gov/sasweb/ncipc/mortrate10_us.html</sup>

As we are more interested in the homicide rate trend, the plot is smoothened to see the clear declining trend of homicide rates.

#### Actual plot (1981-2016 -Smoothened):

```{r Actual DClaim 1 cont..}
ggplot(homicides1) +geom_smooth(aes(Year, norm,color=Homicide_type), se=FALSE)+ggtitle('Firearm and Non-Firearm homicide rates have declined over the years')+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black"), plot.title = element_text( size=10),axis.title= element_text( size=8),legend.title= element_text( size=9))+ labs(x='Year',y = "Number of homicides per 100,000 people")+ scale_x_continuous(breaks=seq(1981,2017,3))+coord_fixed(ratio = 4) +guides(col=guide_legend(title="Homicide type"))
```

<sup>Source: Centers for Disease Control and Prevention-https://webappa.cdc.gov/sasweb/ncipc/mortrate10_us.html</sup>

To explain the importance of gun violence in the US, it is better, if we show the audience that the gun violence has increased over the years instead of showing the opposite. 

Hence, I replicated the plot and it showed the same trend as in the data product. To make the main argument more compelling, I decided to pick up just the increasing trend. But there is no increasing trend in homicides (except during 2015 and 2016, which is very minimal). 

So, I decided to limit the time period which has a constant trend (neither increasing nor decreasing) of homicide rate. It would have been a lot better if we have shown increasing trend of homicide rates. Unfortunately, we don't have such data.

Note that, the homicides does not include deaths inflicted by the police or other law-enforcing agents in the course of arresting or attempting to arrest lawbreakers.

In all the plots below, related to the claim, contrasting colors are used just to show a difference between firearm and non-firearm homicides. The colors, themselves, do not convey any meaning. 


#### Final Version Deceptive plot:

The first and final version of deceptive plot doesn't have much difference except the color of the plot, axis and label font (to make it uniform across all plots). Hence, didn't include the first version again.

```{r Final Version DClaim 1}
homcides_cut=homicides1[homicides1$Year>2002,]

ggplot(homcides_cut) +geom_line(aes(Year, norm,color=Homicide_type ))+ggtitle('Non-firearm homicide rates have declined over years while firearm homicide rates have not')+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),  axis.line = element_line(colour = "black"), plot.title = element_text( size=10),axis.title= element_text( size=9),legend.title= element_text( size=9))+ labs(x='Year',y = "Number of homicides per 100,000 people")+ scale_x_continuous(breaks=seq(2003,2017,1))+coord_fixed(ratio = 2) +guides(col=guide_legend(title="Homicide type"))
```

<sup> Source: Centers for Disease Control and Prevention- -https://webappa.cdc.gov/sasweb/ncipc/mortrate10_us.html</sup>

The plot is smoothened below to see clearly that there is no increasing/decreasing homicide trend.

#### Deceptive plot(Smoothened):

```{r Final Version DClaim 1 cont..}
ggplot(homcides_cut) +geom_smooth(aes(Year, norm,color=Homicide_type), se=FALSE)+ ggtitle('Non-firearm homicide rates have declined over years while firearm homicide rates have not')+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(), axis.line = element_line(colour = "black"), plot.title = element_text( size=10),axis.title= element_text( size=8), legend.title= element_text( size=9))+ labs(x='Year',y = "Number of homicides per 100,000 people")+ scale_x_continuous(breaks=seq(2003,2017,1))+ coord_fixed(ratio = 2)+ guides(col=guide_legend(title="Homicide type"))
```

<sup>Source: Centers for Disease Control and Prevention- -https://webappa.cdc.gov/sasweb/ncipc/mortrate10_us.html</sup>


### Deceptive Claim 2: Tighter gun control laws does not reduce the number of gun-related deaths.

#### Chart from Vox:
```{r Original DClaim 2}
img <- readPNG( '/Users/preethiranganathan/Documents/Santa Clara/data viz/indiv proj/Image8.png',native=TRUE)
plot.new() 
rasterImage(img,0,0,1,1)
```

<sup>Source: Vox https://www.vox.com/policy-and-politics/2017/10/2/16399418/us-gun-violence-statistics-maps-charts</sup>

This original chart is a geographical map. But it is not a good idea to have this map as a warrant for expressing number of deaths. To validate this argument, we just need to focus on states with tighter gun control laws and states with less gun control laws. In this chart, tighter gun control laws are defined as having atleast one firearm law designed to protect children in the state. This might not be a real indicator of strictness of gun control laws. Number of gun control laws that are exercised in a state would be a good metric to measure the strictness of gun control laws. 

#### Actual Argument: 
States with tighter gun control laws have fewer gun-related deaths. 

#### Actual plot:

```{r Actual DClaim 2}
# Gun control laws for each state and year
gun_control=read.csv('~/Documents/Santa Clara/data viz/indiv proj/raw_data.csv',header=TRUE) 

# Gun control for 2014
gun_control=gun_control[gun_control[,2]==2014,]
gun_control=data.frame(gun_control[,1],gun_control[,136])
colnames(gun_control)<-c('State','Number_laws')

# Statewise gun death data -2014
gundeaths2014=read.xls('~/Documents/Santa Clara/data viz/indiv proj/Gun_Deaths_statewise.xlsx',header=TRUE, sheet=2)

# Filtering states with high number of gun control laws
highlaws=data.frame(gun_control[order(-gun_control$Number_laws),][1:10,])
lesslaws=data.frame(gun_control[order(gun_control$Number_laws),][1:10,])
highlaws$Number_GunControl_Laws='high'
lesslaws$Number_GunControl_Laws='low'
highlow=rbind(highlaws,lesslaws)

# Merging data sets
gun_control_deaths=merge(gundeaths2014, highlow ,by.x='STATE.NAME',by.y='State')
guncontrol_deaths=data.frame(gun_control_deaths[,2], gun_control_deaths[,3], gun_control_deaths[,4], gun_control_deaths[,6], gun_control_deaths[,10])
colnames(guncontrol_deaths)<-c("State",'population','Deathrate','Deaths_num','Number_of_gun_control_laws')
guncontrol_deaths$population<-as.numeric(str_replace_all(guncontrol_deaths$population,",",""))

ggplot(guncontrol_deaths, aes(reorder(State, -Deathrate), Deathrate , fill=Number_of_gun_control_laws)) +geom_col(position ='dodge') +ggtitle('States with tighter gun control laws have fewer firearm deaths')+theme_minimal()+theme(plot.title = element_text( size=10),axis.title= element_text( size=9),legend.title= element_text( size=9))+ labs(x='State',y = "Death rate in 2014")+ scale_fill_gradient(high='#00BFC4')+ guides(fill=guide_legend(title="Number of firearm control laws"))
```

<sup>Sources: Centers for Disease Control and Prevention- https://webappa.cdc.gov/sasweb/ncipc/mortrate10_us.html ,
<br />State Firearm laws- https://www.statefirearmlaws.org/table.html</sup>

Here, I decided to go with scale gradient bar chart(continuous colour scales) as the metric chosen is the number of gun laws in a state. So, it is good to show a comparison of deaths for states with diifferent number of gun laws.

Note that, I have retained the grid in the charts for this claim, since in bar charts, it is easier to draw comparison with grids.

#### Deceptive Plot:

This is one of the main claims made by the data product. So, I picked this chart and made a deceptive version out of it.

I picked the states with high gun control laws and states with less gun control laws to see a clear difference in gun deaths between states with high gun control laws and low gun control laws. Here, I normalized the number of deaths in each state per 100,000 people so that the population of each state is taken into consideration. I also noticed that death rate is not dependent on population in this chart. 

Hence, I decided to make a deceptive chart without normalizing for population ie.,I made y axis as raw number of deaths, not death rate per 100,000 people. As I expected, the chart showed that states with high gun control laws has no correlation with gun deaths.

It would have been better, if I had picked few states where gun deaths are prevalent and plotted a time trend of change in number of control laws and change in gun deaths over years. There could be numerous factors in each state, which could cause an increase in gun deaths. By comparing gun deaths of one state across different years, we control for other factors causing gun deaths in different states. 

Unfortunately, that kind of time series data for every state is not available. Hence, I have considered statewise number of gun control laws and gun deaths in 2014.

```{r DClaim 2}
ggplot(guncontrol_deaths, aes(reorder(State, -Deaths_num), Deaths_num , fill=Number_of_gun_control_laws)) + geom_col(position ='dodge') + ggtitle('Tighter gun control laws does not reduce the number of firearm deaths')+ theme_minimal()+ theme(plot.title = element_text(size=10), axis.title= element_text( size=9), legend.title= element_text(size=9))+ labs(x='State',y = "Number of deaths in 2014")+ scale_fill_gradient(high='#00BFC4')+ guides(fill=guide_legend(title="Number of firearm control laws"))
```

<sup>Sources: Centers for Disease Control and Prevention- https://webappa.cdc.gov/sasweb/ncipc/mortrate10_us.html ,
<br />State Firearm laws- https://www.statefirearmlaws.org/table.html</sup>

I have just changed the color of the bars to the uniform color used throughout the chart, added the source and removed the background from the first version. These are the only changes to the chart. Hence, not including the first version.

### Deceptive Claim 3: Restricting access to firearms can save lives.

As the number of firearm control laws has increased in the US over years, firearm violence has reduced. New legal restrictions on owning and purchasing firearms led to a drop in gun violence. 

This argument is from the data product, but this is also deceptive.

#### Actual Argument:
The actual argument (suggested by me) is that restricting access to guns has not saved lives. Gun violence has increased as the number of gun control laws has increased over years. 

#### Actual plot:

The entire data product is just focussing on the homicide rates or suicides rates seperately, sometimes both. But, it never considered the number of gun injuries. So, I calculated the total firearam casualties as a sum of gun homicides, suicides and non-fatal injuries.

```{r Actual DClaim 3,message=FALSE, warning=FALSE}
# Non-fatal gun injury
gun_injuryfiles <- list.files(path='~/Documents/Santa Clara/data viz/indiv proj/firearm_injuries',full.names = TRUE, pattern = ".csv")
d=data.frame()
for(i in 1:length(gun_injuryfiles)){
  d=rbind(read.table(gun_injuryfiles[i],sep=',',header=TRUE),d)
}
gun_injury=data.frame(d[,4],d[,7])
colnames(gun_injury)<-c('Year','Number_injuries')
gun_injury=gun_injury[gun_injury$Year>2000&gun_injury$Year<2015,]
gun_injury$Number_injuries=as.integer(str_replace_all(unfactor(gun_injury$Number_injuries),',',''))

# Gun homicides
firearmhomicidefiles <- list.files(path='~/Documents/Santa Clara/data viz/indiv proj/homicide',full.names = TRUE, pattern = ".csv")
d=data.frame()
for(i in 1:length(firearmhomicidefiles)){
  d=rbind(read.table(firearmhomicidefiles[i],sep=',',header=TRUE),d)
}
firearm_homicide=data.frame(d[,6],d[,9],d[,10])
firearm_homicide=unique(firearm_homicide)
colnames(firearm_homicide)<- c('Year','Homicides','Population')
firearm_homicide=firearm_homicide[firearm_homicide$Year>2000&gun_injury$Year<2015,]

# Gun suicides
suicidefiles <- list.files(path='~/Documents/Santa Clara/data viz/indiv proj/firearm_suicide',full.names = TRUE, pattern = ".csv")
d=data.frame()
for(i in 1:length(suicidefiles)){
  d=rbind(read.table(suicidefiles[i],sep=',',header=TRUE),d)
}
suicide=data.frame(d[,6],d[,9])
suicide=unique(suicide)
colnames(suicide)<- c('Year','Suicides')
suicide=suicide[suicide$Year>2000&gun_injury$Year<2015,]

## Total firearam casualties as a sum of gun homicides, suicides and non-fatal injuries
gun_casualties<-merge(firearm_homicide,suicide,by.x='Year',by.y='Year')
gun_casualties<-merge(gun_casualties,gun_injury,by.x='Year',by.y='Year')
gun_casualties$Total_casualties<-gun_casualties$Homicides+gun_casualties$Suicides+gun_casualties$Number_injuries
gun_casualties$Total_casualties_rate<-(gun_casualties$Total_casualties/gun_casualties$Population)*100000
gun_casualties$Homicide_rate<-(gun_casualties$Homicides)*100000/gun_casualties$Population 

#Gun control laws vs Gun Violence over years
gun_control=read.csv('~/Documents/Santa Clara/data viz/indiv proj/raw_data.csv',header=TRUE) 
gun_control=data.frame(gun_control[,2],gun_control[,136])
colnames(gun_control)<-c('Year','Number_laws')
gun_control=gun_control %>% group_by(Year) %>% summarize(Number_laws=sum(Number_laws))

homicides_control=merge(gun_casualties, gun_control ,by.x='Year',by.y='Year',all=TRUE)
homicides_control=homicides_control[complete.cases(homicides_control),]


p10=ggplot(homicides_control) +geom_line(aes(Year, Total_casualties_rate ))+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),axis.line = element_line(colour = "black"),axis.title= element_text( size=9))+ labs(x='Year',y = "# of deaths per 100,000 people")+ scale_x_continuous(breaks=seq(2001,2014,1)) +coord_fixed(ratio = 0.515)

p11=ggplot(homicides_control) +geom_line(aes(Year, Number_laws ))+ggtitle(' Restricting access to firearms does not help reduce gun violence in the United States')+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),axis.line = element_line(colour = "black"),plot.title = element_text( size=11),axis.title= element_text(size=9))+ labs(x='Year',y = "# of gun laws") + scale_x_continuous(breaks=seq(2001,2014,1))

ggarrange(p11, p10 + font("x.text", size = 10),
                    ncol = 1, nrow = 2)
```

<sup>Sources: Centers for Disease Control and Prevention- https://webappa.cdc.gov/sasweb/ncipc/mortrate10_us.html ,
<br />State Firearm laws- https://www.statefirearmlaws.org/table.html</sup>


This plot clearly shows that there has been an increasing trend of gun casualties over years. We could see that death rate increases as the number of gun laws increases in the US over years.

#### Deceptive plot:

The deceptive argument presented in the data product did not take into account, the total number of firearm violence as a sum of gun homicides, suicides and non-fatal injuries.

The decrease in homicide rate could be due to improved medical facilities and emergency services in the recent years. https://www.umass.edu/newsoffice/article/research-finds-us-murder-rate-suppressed-improved-emergency-medical-response

This could be verified by importing a Emergency Medical Service (EMS) dataset, and  checking the trend to correlate with decrease in homicide rate. 

```{r Final version DClaim 3}
p9=ggplot(homicides_control) +geom_line(aes(Year, Homicide_rate ))+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),axis.line = element_line(colour = "black"),axis.title= element_text( size=9))+ labs(x='Year',y = "# of deaths per 100,000 people")+ scale_x_continuous(breaks=seq(2001,2014,1))+ coord_fixed(ratio = 3)

p11=ggplot(homicides_control) +geom_line(aes(Year, Number_laws ))+ggtitle('Restricting access to firearms can save lives in the United States')+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank(),axis.line = element_line(colour = "black"),plot.title = element_text( size=11),axis.title= element_text(size=9))+ labs(x='Year',y = "# of gun laws") + scale_x_continuous(breaks=seq(2001,2014,1))

ggarrange(p11, p9 + font("x.text", size = 10),
                    ncol = 1, nrow = 2)
```

<sup>Sources: Centers for Disease Control and Prevention- https://webappa.cdc.gov/sasweb/ncipc/mortrate10_us.html ,<br /> 
State Firearm laws- https://www.statefirearmlaws.org/table.html</sup>

I have just changed the color of the curves to black (as there was no necessity to include colors in it), added the source and change the fonts, from the first version. Otherwise, I made no changes to the chart. Hence, not including the first version.

